Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2018-11-30 papers/histogram/sad.tex

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.3$ and a filling fraction of
$\eta = 0.3$. We also explore various translation scales $\delta_0 = 0.005R$ and
$0.05R$. The simulations explore the
energy space of the systems at reduced temperatures of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves
for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05R$). For the $\eta = 30\%$ and $N = 50$ system, we
explore the choice of $\delta_0$ as a parameter.

{\color{red}
$S_{\max} = 120$ and $S_{\min} = 265$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} - S_{\max} = 145$.
The reduced temperature of $1/3$ is chosen in order to examine the convergence
properties of SAD in a reasonable simulation time.
}

Smaller translation
scales cause the given method to take additional time to explore all
energies. Ideal methods should scale linearly with the $\delta_0$, that
is, a decrease in the translation scale should result in a proportional
increase in convergence time. However, as can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} SAMC methods that rapidly
converged for $\delta = 0.05R$ do not proportionaly converge for a
translation scale of $\delta = 0.005R$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/2$ with
  SAMC $t_0 = 10^6$ serving as the reference.}\label{fig:n500-ff0.3}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N500-lndos-comparison}
  \caption{Use this to create description of the system.}\label{fig:n500-lndos-comparison}
\end{figure}

@@

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\eta = 0.3$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly
inserted into the system and random moves are proposed (with a
translation scale $\delta_0 = 0.05R$).

@@

For the larger system, SAMC parameter values of $t_0$ that gave rapid
convergence for the $N = 50$ system with $\delta = 0.05R$ do not
converge rapidly.  Several additional larger $t_0$ parameters are
chosen to compare with SAD.  SAD has significantly smaller fluctuations
in the average entropy error for a given number of moves.

>>>>

For the square-well fluid, our simulations treat systems with a
well-width of $\lambda = 1.3$ and a filling fraction of
$\eta = 0.3$. We also explore various translation scales $\delta_0 = 0.005\sigma$ and
$0.05\sigma$. The simulations explore the
energy space of the systems at reduced temperatures of
$T_{\text{min}} = 0.2$ and $0.5$.  All simulations lead to the minimum important
energy $E_{\min}$ and maximum entropy energy $E_{\max}$
being calculated (with the exception of the WL methods where both
of these parameters are needed a priori).  In the sections below, we
examine the systems parameters impact on the convergence of the methods.

@@

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average entropy vs moves
for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$). For the $\eta = 30\%$ and $N = 50$ system, we
explore the choice of $\delta_0$ as a parameter.

{\color{red}
$S_{\max} = 120$ and $S_{\min} = 248$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} - S_{\max} = 145$.
The reduced temperature of $1/3$ is chosen in order to examine the convergence
properties of SAD in a reasonable simulation time.

As a side note, WL-TMMC does not converge well for $\delta_0 = 0.05\sigma$ while
it does perform well for $\delta_0 = 0.005\sigma$. The WL portion of the algorithm
spends nearly 100x longer for the smaller translation scale ensuring that the algorithm
converges quickly upon transitioning to TMMC.
}

Smaller translation
scales cause the given method to take additional time to explore all
energies. Ideal methods should scale linearly with the $\delta_0$, that
is, a decrease in the translation scale should result in a proportional
increase in convergence time. However, as can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} SAMC methods that rapidly
converged for $\delta = 0.05\sigma$ do not proportionaly converge for a
translation scale of $\delta = 0.005\sigma$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_50-ff0_17-N256-entropy-error-defualt.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.17$ and $T_{\min} = 1$ with
  SAD serving as the reference.}\label{fig:n256}
\end{figure}
\begin{figure}
\includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N500-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/2$ with
  SAMC $t_0 = 10^6$ serving as the reference.}\label{fig:n500-ff0.3}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N500-lndos-comparison}
  \caption{Use this to create description of the system.}\label{fig:n500-lndos-comparison}
\end{figure}

@@

A system is configured with dimensions $x = y = z = 19\sigma$.  A
filling fraction $\eta = 0.3$ is chosen which yields $N = 500$
particles in a cubical lattice system.  These particles are randomly
inserted into the system and random moves are proposed (with a
translation scale $\delta_0 = 0.05\sigma$).

@@

For the larger system, SAMC parameter values of $t_0$ that gave rapid
convergence for the $N = 50$ system with $\delta = 0.05\sigma$ do not
converge rapidly.  Several additional larger $t_0$ parameters are
chosen to compare with SAD.  SAD has significantly smaller fluctuations
in the average entropy error for a given number of moves.
